A fully labelled proof system for intuitionistic modal logics

### 2021-01-15 FullyLabelled/JLC/document.tex
removing some comments
proof read up to end of section 3, will resume later today

\documentclass[a4paper]{article}
\usepackage{authblk}
\input{new-macros}

\title{A fully labelled proof system for intuitionistic modal logics}
\author[1]{Sonia Marin} 
\author[2,3]{Marianela Morales}
\author[3,2]{Lutz Stra{\ss}burger}
\affil[1]{University College London}
\affil[2]{Laboratoire d'Informatique de l'\'Ecole Polytechnique}
\affil[3]{Inria Saclay}

\date{}

@@

 \begin{abstract}
   %% \review{Reconsider abstract. It reads very “operational” and does not clearly state the contributions of the paper.}
   %% \lutz{Done?}
   %% \sonia{Added two sentences at the beginning and modified it a bit.}
   Labelled proof theory has been famously successful for modal logics by mimicking their relational semantics within deductive systems. Simpson in particular designed a framework to study a variety of intuitionistic modal logics integrating a binary relation symbol in the syntax.
   %%
   In this paper, we present a labelled sequent system for
   intuitionistic modal logics such that there is not only one, but
   two relation symbols appearing in sequents: one for the accessibility
   relation associated with the Kripke semantics for normal modal logics and
   one for the preorder relation associated with the Kripke semantics
   for intuitionistic logic.
   %%
   This puts our system in close correspondence with the standard
   birelational Kripke semantics for intuitionistic modal logics. 
   %%
   As a consequence it can be extended with arbitrary intuitionistic Scott-Lemmon axioms. 
   %%
   We show soundness and completeness, together with an internal cut elimination proof, encompassing a wider array of intuitionistic modal logics than any existing labelled system.
 \end{abstract}

@@

 Labelled sequent calculi for logics with
 standard Kripke semantics attach to every formla $\fm A$ a label
 $\lb x$, written as $\labels xA$, and additionally use
 \emph{relational atoms} of the form $\accs xy$ where 
 the binary relation symbol $\rel$
 %These calculi work best for logics with
 %standard Kripke semantics, as in this case the relation $\rel$ 
 is used to encode the accessibility relation in the Kripke models, such that
 frame conditions corresponding to the desired logic can be directly
 encoded as inference rules.
 %\todo{Reviewer 1 complained about the ``best''}
 Prominent examples are classical modal
 logics and intuitionistic propositional logic, e.g., the frame
 condition of transitivity ($\forall \lb {xyz}.\:\accs xy\AND \accs yz\IMP\accs xz$) can be translated into the
 inference rule
 \begin{equation}
 \label{eq:Rtrans}
 \vlinf{\rn{trans}}{}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
 	\B, \accs xy, \accs yz, \accs xz, \Left \SEQ \Right}
 \end{equation}
 where $\B$ stands for a set of relational atoms, and $\Gamma$ and
 $\Delta$ for multi-sets of labelled formulas~\cite{negri:jpl2005}.

 However,
 in this paper we are concerned with \emph{intuitionistic modal
 	logics} whose Kripke semantics is based on \emph{birelational}
 frames, i.e., they have two binary relations instead of one: one
 relation $R$ that corresponds to the accessibility relation in 
 Kripke frames for modal logics, and a relation $\le$ that
 corresponds to the preorder relation in Kripke frames for
 intuitionistic logic. Consequently, standard labelled systems for
 these logics have certain shortcomings:
 \begin{enumerate}
 	\item The transitivity rule in~\eqref{eq:Rtrans} can be axiomatised by the conjunction of the two
 	versions of the $\fourax$-axiom
 	\begin{equation}
 	\label{eq:4ax}
 	\ax{4^{\BOX}}\colon\fm{\BOX A\IMP \BOX\BOX A}
 	\qquand
 	\rn{4^{\DIA}}\colon\fm{\DIA\DIA A\IMP\DIA A}
 	%	\quadcm
 	\end{equation}
 	which are equivalent in classical
 	modal logic. However, in intuitionistic modal logic they are not
 	equivalent, and even though the logic $\IKfour$ 
 	(intuitionistic version of modal logic $\Kfour$) contains both
 	axioms, they can also be added independently to the logic $\IK$ 
 	(intuitionistic version of $\K$). The proof theory of these
 	distinct logics has not been discussed before, but their
 	corresponding frame conditions
 	\begin{equation}
 	\label{eq:4frame}
 	\forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{x'}.\:\futs x{x'}\AND\accs {x'}z)
 	\quand
 	\forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{z'}.\:\futs z{z'}\AND\accs {x}{z'})
 	\;
 	\end{equation}
 	%  \todo{add graphical representation of the conditions}
 	$$
 	\xymatrix{
 		x' \ar@{.>}[drr]^R \\
 		x \ar@{.>}[u]^\le \ar@{->}[r]_{R} & y \ar@{->}[r]_{R} & z
 	}
 	\hskip10em
 	\xymatrix{
 		&& z''  \\
 		x  \ar@{.>}[urr]^R \ar@{->}[r]_{R} & y \ar@{->}[r]_{R} & z \ar@{.>}[u]^\le
 	}
 	$$
 	respectively, have been studied in~\cite{plotkin:stirling:86}.
 	%  \sonia{the general condition is correct but not the application to 4}
 	\item The correspondence between the syntax and the semantics is not
 	as clean as one would expect. As only the $R$-relation (and not the
 	$\le$-relation) is visible in an ordinary labelled
 	sequent, we only have the result that a formula is provable in an extension of $\IK$ beyond paths axioms 
 	if and only if is satisfied in all \emph{graph-consistent}\footnote{This
 		means that every layer in the model can be lifted to any future of
 		any world in that layer. See~\cite{simpson:phd} and~\cite{mar:str:tableaux17} for a formal definition and discussion.}   
 	models satisfying the corresponding conditions, as observed by Simpson	and considered an inelegant solution~\cite{simpson:phd}.
 \end{enumerate}
 In order to address these two concerns we follow here the approach~\cite{garg2012,maffezioli:etal:synthese13} that enriches usual  labelled sequents by allowing both, relational atoms of the form $\futs xy$ and of the form $\accs xy$ and develop it into a uniform general framework. 
 %%
 Consequently, following the same ``axiom-to-rule'' method as Simpson, we can translate the frame
 conditions in~\eqref{eq:4frame} into inference rules:
 \begin{equation}
 \label{eq:4rules}
 \vlinf{\rn{4^{\BOX}}}{\proviso{$\lb{x'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
 	\B, \accs xy, \accs yz, \accs {x'}z, \futs x{x'},\Left \SEQ \Right}
 \qquand
 \vlinf{\rn{4^{\DIA}}}{\proviso{$\lb{z'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
 	\B, \accs xy, \accs yz, \accs {x}{z'}, \futs z{z'},\Left \SEQ \Right}  
 \end{equation}
 This allows us to define cut-free deductive systems for a wide range
 of logics that had not been treated proof-theoretically before.  
 %%
 Furthermore, the relation
 between syntax and semantics can be recovered as one would expect: A formula is
 provable in this system if and only if it is valid in all models satisfying the corresponding frame condition.

 Besides that, there is another pleasant observation to make about our
 system. It is naturally a multi-conclusion system. Ordinary labelled sequent systems for intuitionistic modal
 logic are single-conclusion~\cite{simpson:phd}. 
 %
 The same is true for the corresponding nested sequent
 systems~\cite{str:fossacs13,marin:str:aiml}. It is possible to express
 Maehara style multi-conclusion systems in nested
 sequents~\cite{str:2017maehara}, and therefore also in ordinary
 labelled sequents. However, even in these systems there are rules
 ($\rlabrn\IMP$ and $\rlabrn\BOX$) that force a single-conclusion
 premise, even though this is not the case in labelled
 systems~\cite{negri:jpl2005} or nested sequents~\cite{fitting:2014} for
 intuitionistic logic.

@@

 This paper is organized as follows. In the next section (Section~\ref{sec:intmod}) we recall the standard syntax and semantics of intuitionistic modal logics. Then, in Section~\ref{sec:system} we present our system for the intuitionistic modal logic $\IK$. In Sections~\ref{sec:completeness} and~\ref{sec:soundness}, we show soundness and completeness of the system with cut. The cut elimination theorem, proved in Section~\ref{sec:cut-elim}, then entails soundness and completeness for the cut-free system. Finally, in Section~\ref{sec:ext} we discuss the possible extension to the system to capture intuitionistic Scott-Lemmon modal logics.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %
 %
 \section{Intuitionistic modal logics}\label{sec:intmod}

@@

 Obtaining the intuitionistic variant of $\mathsf{K}$ is more involved than the classical variant. 
 %%
 Lacking De Morgan duality, there are many variants of the \emph{distributivity axiom} $\mathsf{k}$ that are classically but not intuitionistically equivalent. Five axioms have been considered as primitives in the literature. An intuitionistic variant of the modal logic $\K$ can then be obtained from ordinary intuitionistic propositional logic $\rn{IPL}$ by adding:
 %

 %The axiomatisation that is now generally accepted as intuitionistic modal logic $\IK$ was given by Plotkin and Stirling~\cite{Plotkin} and is equivalent to the one proposed by Fischer-Servi~\cite{Fischer}. It is obtained from intuitionistic propositional logic $\mathsf{IPL}$ by adding:
 \begin{itemize}
 	\item the \emph{necessitation rule}: if $\fm A$ is a theorem then $\fm{\BOX A}$ is also a theorem; and
 	\item the following five variants of $\rn{k}$:
 	\begin{equation*}
 		\label{eq:ik}%\hskip-2em
 		\begin{array}[t]{r@{\;}l@{\qquad}r@{\;}l@{\qquad}r@{\;}l}
 			\kax[1]\colon&\fm{\BOX(A\IMP B)\IMP(\BOX A\IMP\BOX B)}
 			&
 			\kax[3]\colon&\fm{\DIA(A\OR B)\IMP(\DIA A\OR\DIA B)}
 			&
 			\kax[5]\colon&\fm{\DIA\BOT\IMP\BOT}
 			\\
 			\kax[2]\colon&\fm{\BOX(A\IMP B)\IMP(\DIA A\IMP\DIA B)}
 			&
 			\kax[4]\colon&\fm{(\DIA A\IMP \BOX B)\IMP\BOX(A\IMP B)}\\%x[1ex]
 		\end{array}
 	\end{equation*}
 \end{itemize}

@@

 The Kripke semantics for $\IK$ was first defined by Fischer-Servi~\cite{fischer-servi:84}.
 %
 It combines the semantics for intuitionistic propositional logic and the one for classical modal logic, using two distinct relations on the set of worlds.\footnote{This semantics and some variants have also been studied in~\cite{bovzic1984models}. For an excellent survey on possible variants see~\cite{simpson:phd}.}

>>>>

\documentclass[a4paper]{article}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{authblk}
\usepackage[margin=2cm]{geometry}

\usepackage{amsmath} 
%\usepackage{amssymb} 
\usepackage{amsthm}

\usepackage{latexsym} % For \Box and \Diamond
\usepackage{colonequals} % for ::=
%\usepackage{bm} % nice boldface for maths

\usepackage[matrix,arrow]{xy}
%\usepackage{xcolor}
\usepackage[noxy]{virginialake}

%\usepackage{array}
\usepackage{graphicx} % For command \includegraphics and \scalebox                        
%\usepackage{tikz}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Small equation environments
\newdimen\mydisplayskip
\mydisplayskip=.4\abovedisplayskip
%\newenvironment{smallequation}
%{\par\nobreak\vskip\mydisplayskip\noindent\bgroup\small\csname equation\endcsname}{\csname endequation\endcsname\egroup}
%%
\newenvironment{smallequation*}
{\par\nobreak\vskip\mydisplayskip\noindent\bgroup\small\csname equation*\endcsname}{\csname endequation*\endcsname\egroup}
%%
%\newenvironment{smallalign}
%{\par\nobreak\noindent\bgroup\small\csname align\endcsname}{\csname endalign\endcsname\egroup}
%%
%\newenvironment{smallalign*}
%{\par\nobreak\noindent\bgroup\small\csname align*\endcsname}{\csname endalign*\endcsname\egroup}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Theorem counter
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
%\newtheorem{conjecture}[theorem]{Conjecture}
%\newtheorem{fact}[theorem]{Fact}
%\newtheorem{claim}[theorem]{Claim}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
%\newtheorem{corollary}[theorem]{Corollary}
%%
\theoremstyle{definition}
%\newtheorem{observation}[theorem]{Observation}
\newtheorem{definition}[theorem]{Definition}
%\newtheorem{problem}[theorem]{Problem}
%\newtheorem{construction}[theorem]{Construction}
\newtheorem{remark}[theorem]{Remark}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Virginialake add-ons
\newcommand{\vlderivationauxnc}[1]{#1{\box\derboxone}\vlderivationterm}
\newcommand{\vlderivationnc}{\vlderivationinit\vlderivationauxnc}
%
%
\makeatletter
\newbox\@conclbox
\newdimen\@conclheight
%
%
\newcommand{\vlhtr}[2]{\vlpd{#1}{}{#2}}
\newcommand\vlderiibase[5]{{%
		\setbox\@conclbox=\hbox{$#3$}\relax%
		\@conclheight=\ht\@conclbox%
		\setbox\@conclbox=\hbox{$%
			\vlderivationnc{%
				\vliin{#1}{#2}{\box\@conclbox}{#4}{#5}%
			}$}%
		\lower\@conclheight\box\@conclbox%
}}
%
\newcommand\vlderibase[4]{{%
		\setbox\@conclbox=\hbox{$#3$}\relax%
		\@conclheight=\ht\@conclbox%
		\setbox\@conclbox=\hbox{$%
			\vlderivationnc{%
				\vlin{#1}{#2}{\box\@conclbox}{#4}%
			}$}%
		\lower\@conclheight\box\@conclbox%
}}
%
\newcommand\vlderidbase[4]{{%
		\setbox\@conclbox=\hbox{$#3$}\relax%
		\@conclheight=\ht\@conclbox%
		\setbox\@conclbox=\hbox{$%
			\vlderivationnc{%
				\vlid{#1}{#2}{\box\@conclbox}{#4}%
			}$}%
		\lower\@conclheight\box\@conclbox%
}}
%
\makeatother
%\left(

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% MACROS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%% Comments  
%\definecolor{notgray}{rgb}{.8,.8,.8}
%\newcommand{\marianela}[1]{{\color{purple}[Marianela: #1]}}
%\newcommand{\review}[1]{{\color{gray}[Reviewer: #1]}}
%\newcommand{\sonia}[1]{{\color{blue}[Sonia: #1]}}
%\newcommand{\lutz}[1]{{\color{notgreen}[Lutz: #1]}}
%\newcommand{\todo}[1]{{\color{red}[TODO: #1]}}
%
%\newcommand{\old}[1]{{\color{notgray}[OLD: #1]}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% General
\newcommand*{\A}{\mathcal{A}}
%\newcommand{\G}{\mathcal{G}}
\newcommand*{\SG}{\fm{\mathcal{G}}}
\newcommand{\SGi}[1]{\fm{\mathcal{G}_{#1}}}
%
\newcommand{\quand}{\quad\mbox{and}\quad}
\newcommand{\qquand}{\qquad\mbox{and}\qquad}
\newcommand{\quadcm}{\rlap{\quad,}}
%
\newcommand{\proviso}[1]{\mbox{\scriptsize #1}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Systems
\newcommand*{\ax}[1]{\mathsf{#1}}
\newcommand*{\kax}[1][]{\ax{k_{#1}}}
\newcommand{\fourax}{\ax{4}}
\newcommand{\agklmn}{{g_{klmn}}}
%
\newcommand*{\IK}{\mathsf{IK}}
\newcommand*{\K}{\mathsf{K}}
%
\newcommand*{\IKfour}{\mathsf{IK4}}
\newcommand*{\Kfour}{\mathsf{K4}}
%
%\newcommand*{\ISfour}{\mathsf{IS4}}
%\newcommand*{\Sfour}{\mathsf{S4}}
%
\newcommand*{\labIKp}{\lab\IK_{\le}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Connectives
\newcommand*{\NOT}{\neg}
\newcommand*{\AND}{\mathbin{\wedge}}
\newcommand*{\TOP}{\mathord{\top}}
\newcommand*{\OR}{\mathbin{\vee}}
\newcommand*{\BOT}{\mathord{\bot}}
\newcommand*{\IMP}{\mathbin{\supset}}%%\scalebox{.9}{\raise.2ex\hbox{$\supset$}}}}

\newcommand*{\BOX}{\mathord{\Box}}
\newcommand*{\DIA}{\mathord{\Diamond}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Labelled sequents
\definecolor{notgreen}{rgb}{.1,.6,.1}

\newcommand{\lseq}[3]{#1 , #2 \SEQ #3}
%
\newcommand{\B}{\mathcal{R}}
\newcommand{\Left}{\Gamma} %{\mathcal{L}}
\newcommand{\Right}{\Delta} %{\mathcal{R}}
%
\newcommand*{\fm}[1]{{\color{notgreen}{#1}}}
\newcommand*{\lb}[1]{{\color{blue}{#1}}}
%
\newcommand*{\rel}{R}
\newcommand*{\labels}[2]{\lb{#1}\mathord{:}\fm{#2}}
\newcommand*{\accs}[2]{\lb{#1}R\lb{#2}}
\newcommand*{\futs}[2]{\lb{#1}\le{\lb{#2}}}
%
\newcommand{\SEQ}{\Longrightarrow}
%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Labelled rules
\newcommand*{\rn}[1]  {\ensuremath{\mathsf{#1}}}
\newcommand*{\lab}{\mathsf{lab}}
%
\newcommand*{\labrn}[2][]  {\rn{#2}_{#1}}%^{\lab}}}
\newcommand*{\rlabrn}[2][]  {\rn{#2}_\rn{R#1}}%^\lab}}
\newcommand*{\llabrn}[2][]  {\rn{#2}_\rn{L#1}}%^\lab}}
%%
%\newcommand*{\brsym}{\boxtimes}%\mathord{\scalebox{.8}{$\blacksquare$}}}
\newcommand*{\diasym}{\diamondplus}%\mathord{\blacklozenge}}
%%
\newcommand*{\boxbrn}[1]{\boxtimes_\rn{#1}}%^{\lab}}}
\newcommand*{\diabrn}[1]{\diamondplus_\rn{#1}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% System labIK+gklmn
\newcommand{\gklmn}{{\mathsf{g_{klmn}}}}%%\boxtimes
\newcommand{\boxk}{\square_{R}^{k}}
\newcommand{\boxlk}{\square_{L}^{k}}
\newcommand{\diamk}{\lozenge_{L}^{k}}
\newcommand{\diamrk}{\lozenge_{R}^{k}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Semantics
\newcommand{\f}{f^{\mathcal{M}}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\F}{\mathcal{F}}
%
\newcommand{\inter}[1]{\lb{\llbracket #1\rrbracket}}
%\newcommand{\force}[3]{#1,#2\Vdash#3}
\newcommand{\nforce}[3]{#1,#2\not\Vdash#3}
\newcommand{\cforce}[3]{#1,\lb{#2}\Vdash\fm{#3}}
\newcommand{\cnforce}[3]{#1,\lb{#2}\not\Vdash\fm{#3}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Derivations
\newcommand{\Dw}{\mathcal{D}^{\rn w}}
\newcommand{\Dwone}{\mathcal{D}_{1}^{\rn w}}
\newcommand{\Dwtwo}{\mathcal{D}_{2}^{\rn w}}
%
\newcommand{\D}{\mathcal{D}}
\newcommand*{\DD}{\mathcal{D}}
\newcommand{\Done}{\mathcal{D}_{1}}
\newcommand{\Dtwo}{\mathcal{D}_{2}}
%
\newcommand{\height}[1]{|#1|}
%
\newcommand*{\reducesto}{\quad{\leadsto}\quad}
\newcommand*{\invr}[1]{#1}%^\bullet}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{A fully labelled proof system for intuitionistic modal logics}
\author[1]{Sonia Marin} 
\author[2,3]{Marianela Morales}
\author[3,2]{Lutz Stra{\ss}burger}
\affil[1]{University College London}
\affil[2]{Laboratoire d'Informatique de l'\'Ecole Polytechnique}
\affil[3]{Inria Saclay}
\date{}

@@

 \begin{abstract}
   Labelled proof theory has been famously successful for modal logics by mimicking their relational semantics within deductive systems. Simpson in particular designed a framework to study a variety of intuitionistic modal logics integrating a binary relation symbol in the syntax.
   %%
   In this paper, we present a labelled sequent system for
   intuitionistic modal logics such that there is not only one, but
   two relation symbols appearing in sequents: one for the accessibility
   relation associated with the Kripke semantics for normal modal logics and
   one for the preorder relation associated with the Kripke semantics
   for intuitionistic logic.
   %%
   This puts our system in close correspondence with the standard
   birelational Kripke semantics for intuitionistic modal logics. 
   %%
   As a consequence it can be extended with arbitrary intuitionistic Scott-Lemmon axioms. 
   %%
   We show soundness and completeness, together with an internal cut elimination proof, encompassing a wider array of intuitionistic modal logics than any existing labelled system.
 \end{abstract}

@@

 Labelled sequent calculi for logics with
 standard Kripke semantics attach to every formla $\fm A$ a label
 $\lb x$, written as $\labels xA$, and additionally use
 \emph{relational atoms} of the form $\accs xy$ where 
 the binary relation symbol $\rel$
 %
 is used to encode the accessibility relation in the Kripke models, such that
 frame conditions corresponding to the desired logic can be directly
 encoded as inference rules.
 %
 Prominent examples are classical modal
 logics and intuitionistic propositional logic, e.g., the frame
 condition of transitivity ($\forall \lb {xyz}.\:\accs xy\AND \accs yz\IMP\accs xz$) can be translated into the
 inference rule
 \begin{equation}
 \label{eq:Rtrans}
 \vlinf{\rn{trans}}{}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
 	\B, \accs xy, \accs yz, \accs xz, \Left \SEQ \Right}
 \end{equation}
 where $\B$ stands for a set of relational atoms, and $\Gamma$ and
 $\Delta$ for multi-sets of labelled formulas~\cite{negri:jpl2005}.

 However,
 in this paper we are concerned with \emph{intuitionistic modal
 	logics} whose Kripke semantics is based on \emph{birelational}
 frames, i.e., they have two binary relations instead of one: one
 relation $R$ that corresponds to the accessibility relation in 
 Kripke frames for modal logics, and a relation $\le$ that
 corresponds to the preorder relation in Kripke frames for
 intuitionistic logic. Consequently, standard labelled systems for
 these logics have certain shortcomings:
 \begin{enumerate}
 	\item The transitivity rule in~\eqref{eq:Rtrans} can be axiomatised by the conjunction of the two
 	versions of the $\fourax$-axiom
 	\begin{equation}
 	\label{eq:4ax}
 	\ax{4^{\BOX}}\colon\fm{\BOX A\IMP \BOX\BOX A}
 	\qquand
 	\rn{4^{\DIA}}\colon\fm{\DIA\DIA A\IMP\DIA A}
 	%	\quadcm
 	\end{equation}
 	which are equivalent in classical
 	modal logic. However, in intuitionistic modal logic they are not
 	equivalent, and even though the logic $\IKfour$ 
 	(intuitionistic version of modal logic $\Kfour$) contains both
 	axioms, they can also be added independently to the logic $\IK$ 
 	(intuitionistic version of $\K$). The proof theory of these
 	distinct logics has not been discussed before, but their
 	corresponding frame conditions
 	\begin{equation}
 	\label{eq:4frame}
 	\forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{x'}.\:\futs x{x'}\AND\accs {x'}z)
 	\quand
 	\forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{z'}.\:\futs z{z'}\AND\accs {x}{z'})
 	\;
 	\end{equation}
 	%  
 	$$
 	\xymatrix{
 		x' \ar@{.>}[drr]^R \\
 		x \ar@{.>}[u]^\le \ar@{->}[r]_{R} & y \ar@{->}[r]_{R} & z
 	}
 	\hskip10em
 	\xymatrix{
 		&& z''  \\
 		x  \ar@{.>}[urr]^R \ar@{->}[r]_{R} & y \ar@{->}[r]_{R} & z \ar@{.>}[u]^\le
 	}
 	$$
 	respectively, have been studied in~\cite{plotkin:stirling:86}.
 	%  
 	\item The correspondence between the syntax and the semantics is not
 	as clean as one would expect. As only the $R$-relation (and not the
 	$\le$-relation) is visible in an ordinary labelled
 	sequent, we only have the result that a formula is provable in an extension of $\IK$ \emph{beyond paths axioms }
 	if and only if it is satisfied in all \emph{graph-consistent}\footnote{This
 		means that every layer in the model can be lifted to any future of
 		any world in that layer. See~\cite{simpson:phd} and~\cite{mar:str:tableaux17} for a formal definition and discussion.}   
 	models satisfying the corresponding standard (\emph{classical}) conditions, as observed by Simpson	and considered an inelegant solution~\cite{simpson:phd}.
 \end{enumerate}
 In order to address these two concerns we follow here the approach~\cite{garg2012,maffezioli:etal:synthese13} that enriches usual  labelled sequents by allowing both, relational atoms of the form $\futs xy$ and of the form $\accs xy$ and develop it into a uniform general framework. 
 %%
 Consequently, following the ``axiom-to-rule'' method, we can translate the frame
 conditions in~\eqref{eq:4frame} into inference rules:
 \begin{equation}
 \label{eq:4rules}
 \vlinf{\rn{4^{\BOX}}}{\proviso{$\lb{x'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
 	\B, \accs xy, \accs yz, \accs {x'}z, \futs x{x'},\Left \SEQ \Right}
 \qquand
 \vlinf{\rn{4^{\DIA}}}{\proviso{$\lb{z'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
 	\B, \accs xy, \accs yz, \accs {x}{z'}, \futs z{z'},\Left \SEQ \Right}  
 \end{equation}
 This allows us to define cut-free deductive systems for a wide range
 of logics that had not been treated proof-theoretically before.  
 %%
 Furthermore, the relation
 between syntax and semantics can be recovered as one would expect: A formula is
 provable in this system if and only if it is valid in all models satisfying the corresponding frame condition.

 Besides that, there is another pleasant observation to make about our
 system: It is naturally a multi-conclusion system. 
 Ordinary labelled sequent systems for intuitionistic modal
 logic are single-conclusion~\cite{simpson:phd}. 
 %
 The same is true for the corresponding nested sequent
 systems~\cite{str:fossacs13,marin:str:aiml}. It is possible to express
 Maehara style multi-conclusion systems in nested
 sequents~\cite{str:2017maehara}, and therefore also in ordinary
 labelled sequents. However, even in these systems there are rules
 ($\rlabrn\IMP$ and $\rlabrn\BOX$) that force a single-conclusion
 premise, even though this is not the case in labelled
 systems~\cite{negri:jpl2005} or nested sequents~\cite{fitting:2014} for
 intuitionistic logic.

@@

 This paper is organized as follows. In the next section (Section~\ref{sec:intmod}) we recall the standard syntax and semantics of intuitionistic modal logics. Then, in Section~\ref{sec:system} we present our system for the intuitionistic modal logic $\IK$. In Sections~\ref{sec:completeness} and~\ref{sec:soundness}, we show its soundness and completeness with cut. The cut elimination theorem, proved in Section~\ref{sec:cut-elim}, then entails soundness and completeness for the cut-free system. Finally, in Section~\ref{sec:ext} we discuss the extension of the system to capture intuitionistic Scott-Lemmon modal logics.

 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Intuitionistic modal logics}\label{sec:intmod}

@@

 Obtaining the intuitionistic version of $\mathsf{K}$ is more involved than the classical one. 
 %%
 Lacking De Morgan duality, there are many variants of the \emph{distributivity axiom} $\mathsf{k}$ that are classical but not intuitionistic consequences of one another. Five axioms have been considered as primitives in the literature. An intuitionistic version of modal logic $\K$ can then be obtained from ordinary intuitionistic propositional logic $\rn{IPL}$ by adding:
 %

 \begin{itemize}
 	\item the \emph{necessitation rule}: if $\fm A$ is a theorem then $\fm{\BOX A}$ is also a theorem; and
 	\item the following five variants of $\rn{k}$:
 	\begin{equation*}
 		\label{eq:ik}%\hskip-2em
 		\begin{array}[t]{r@{\;}l@{\qquad}r@{\;}l@{\qquad}r@{\;}l}
 			\kax[1]\colon&\fm{\BOX(A\IMP B)\IMP(\BOX A\IMP\BOX B)}
 			&
 			\kax[3]\colon&\fm{\DIA(A\OR B)\IMP(\DIA A\OR\DIA B)}
 			&
 			\kax[5]\colon&\fm{\DIA\BOT\IMP\BOT}
 			\\
 			\kax[2]\colon&\fm{\BOX(A\IMP B)\IMP(\DIA A\IMP\DIA B)}
 			&
 			\kax[4]\colon&\fm{(\DIA A\IMP \BOX B)\IMP\BOX(A\IMP B)}\\%x[1ex]
 		\end{array}
 	\end{equation*}
 \end{itemize}

@@

 The Kripke semantics for $\IK$ was first defined by Fischer-Servi~\cite{fischer-servi:84}.
 %
 It combines the semantics for intuitionistic propositional logic and the one for classical modal logic, using two distinct relations on the set of worlds.\footnote{This semantics and some variants have also been studied in~\cite{bovzic1984models}. For an excellent survey on possible alternatives see~\cite{simpson:phd}.}
